Diffantom: a software layer to build digital phantoms of whole-brain diffusion MRI

### 2015-07-09 paper/main.tex
add new intro

\linenumbers
\section*{Introduction}
The assessment of tractography from \gls*{dmri} data and subsequent analyses is challenging due to
  the lack of a gold standard.
Large efforts have been devoted to the development of physical phantoms
  \citep{lin_validation_2001,campbell_flowbased_2005,perrin_validation_2005,fieremans_simulation_2008,tournier_resolving_2008}.
\cite{cote_tractometer_2013} conducted a thorough review of tractography methodologies using the
  so-called \emph{FiberCup} phantom \citep{poupon_new_2008,fillard_quantitative_2011}.
These phantoms are appropriate to evaluate the resolution of fiber crossings and accuracy of
  direction-independent scalar parameters in very simplistic geometries.
Since the complexity of whole-brain tractography is unaccountable with current materials and proposed
  methodologies to build physical phantoms, digital simulations are increasingly popular.
Early digital phantoms started with simulation of simple geometries as well
  \citep{basser_in_2000,goessl_fiber_2002,tournier_limitations_2002,leemans_mathematical_2005}
  to evaluate resolution of fiber crossings.
These tools generally implemented the multi-tensor model \citep{alexander_analysis_2001,tuch_high_2002}
  to simulate fibers crossing, fanning, kissing, etc.
\cite{close_software_2009} presented the \emph{Numerical Fibre Generator}, a software to simulate
  spherical shapes filled with digital fiber-tracts.
\cite{caruyer_phantomas_2014} proposed \emph{Phantomas} to simulate any kind of analytic geometry
  inside a sphere.
\emph{Phantomas} models diffusion by a restricted and a hindered compartment, similar to
  \citep{assaf_composite_2005}.
\cite{wilkins_fiber_2014} proposed a whole-brain simulated phantom derived from voxel-wise orientation
  of fibers averaged from real \gls*{dmri} scans and the multi-tensor model with a compartment of
  isotropic-diffusion.
Recently, \cite{neher_fiberfox_2014} proposed \emph{FiberFox}, a visualization software to develop
  complex geometries and their analytical description.
Once the geometries are obtained, the software generates the corresponding \gls*{dmri} signal with a
  methodology very close to that implemented in \emph{Phantomas}.

In this paper we present \emph{Diffantom}, a software layer to generate diffusion phantoms at whole-brain
  scale and derived from real \gls*{dmri} datasets.
Particularly, we propose the \gls*{hcp} \citep{essen_human_2012} as the source of real data.
Finally, an example dataset is openly released along with all the necessary requirements and instructions
  to generate custom phantoms.

>>>>

\linenumbers

\section*{Introduction}
Fiber tracking on \gls*{dmri} data has become an important application in the in-vivo investigation
  of the structural configuration of fiber bundles at the macro-scale.
The final outcome of such applications, called tractography, is fundamental to gain information about
  \gls*{wm} morphology in neurosurgical planning, post-surgery evaluations, and the study of neurological
  diseases such as multiple sclerosis, and Alzheimer's disease \citep{chua_diffusion_2008}.
Tractography is also crucial in the analysis of structural brain networks using graph theory.
These analyses have been successfully applied in the definition of the unique subject-wise patterns of connectivity
  \citep{sporns_human_2005}, the assessment of neurological diseases \citep{griffa_structural_2013}, and in the
  study of the link of structural and functional connectivity \citep{messe_predicting_2015}.
However, the development of the field is limited by the lack of gold standards to test and compare the
  wide range of algorithms and processing methodologies available.
In this work we propose \emph{diffantom}, an \emph{in-silico} dataset to assess tractography pipelines using
  real datasets as source of whole-brain micro-structural information.
We also provide a full software layer to produce \emph{diffantoms} from real \gls*{dmri} datasets
  drawn from the \gls*{hcp}.
This dataset enables the following evaluations of standard connectivity pipelines:
\begin{itemize}
  \item Comparison of the impact of the different diffusion sampling schemes on the local micro-structure
  model of choice and the subsequent global tractography outcome.
  \item Investigation of the effects of the different artifacts (noise, susceptibility-derived, Eddy-currents,
  etc.) on the final connectivity matrix.
  \item Determine of the \acrfull*{roc} of connectivity pipelines.
  \item Generation of synthetic connectivity networks to test graph theory analyses.
\end{itemize}

\section*{Background}
The assessment of tractography from \gls*{dmri} data and subsequent analyses is challenging due to
  the lack of a gold standard.
Large efforts have been devoted to the development of physical phantoms
  \citep{lin_validation_2001,campbell_flowbased_2005,perrin_validation_2005,fieremans_simulation_2008,tournier_resolving_2008}.
\cite{cote_tractometer_2013} conducted a thorough review of tractography methodologies using the
  so-called \emph{FiberCup} phantom \citep{poupon_new_2008,fillard_quantitative_2011}.
These phantoms are appropriate to evaluate the resolution of fiber crossings and accuracy of
  direction-independent scalar parameters in very simplistic geometries.
Since the complexity of whole-brain tractography is unaccountable with current materials and proposed
  methodologies to build physical phantoms, digital simulations are increasingly popular.
Early digital phantoms started with simulation of simple geometries as well
  \citep{basser_in_2000,goessl_fiber_2002,tournier_limitations_2002,leemans_mathematical_2005}
  to evaluate resolution of fiber crossings.
These tools generally implemented the multi-tensor model \citep{alexander_analysis_2001,tuch_high_2002}
  to simulate fibers crossing, fanning, kissing, etc.
\cite{close_software_2009} presented the \emph{Numerical Fibre Generator}, a software to simulate
  spherical shapes filled with digital fiber-tracts.
\cite{caruyer_phantomas_2014} proposed \emph{Phantomas} to simulate any kind of analytic geometry
  inside a sphere.
\emph{Phantomas} models diffusion by a restricted and a hindered compartment, similar to
  \citep{assaf_composite_2005}.
\cite{wilkins_fiber_2014} proposed a whole-brain simulated phantom derived from voxel-wise orientation
  of fibers averaged from real \gls*{dmri} scans and the multi-tensor model with a compartment of
  isotropic-diffusion.
Recently, \cite{neher_fiberfox_2014} proposed \emph{FiberFox}, a visualization software to develop
  complex geometries and their analytical description.
Once the geometries are obtained, the software generates the corresponding \gls*{dmri} signal with a
  methodology very close to that implemented in \emph{Phantomas}.
